Symbolic Abstract Data Type Inference

### 2015-07-09 discussion.tex
some final typos
Removed kernel talk.

First, while our theoretical foundation does cover “concurrency-aware” ADT
specifications~\cite{conf/podc/HemedR14}, i.e.,~non-atomic ADTs like the
\emph{rendezvous synchronizer}, \emph{barrier}, and \emph{exchanger}, whose
kernels have bounded width greater than $1$, our current implementation only
handles atomic ADTs. In principle, this limitation is not fundamental. The key
difference is in the enumeration of kernel histories. For a given ADT,
determining whether a given history is in the kernel or its complement reduces
to checking whether there exists an execution of which the history is an
abstraction. For atomic ADTs, only a single sequential execution need be
examined, in which no two operations overlap. For non-atomic ADTs, every
possible interleaving of the internal actions of operations need be examined.
While we expect this calculation to remain feasible given that ADT complements
are normally represented with histories with few operations, we do expect it to
incur a noticeable cost.

Second, ADTs whose specifications rely on relations (besides equality) on
method argument and return values cannot be expressed with our notion of
histories based on matching schemes. A notable example of such an ADT is the
\emph{priority queue}, whose \emph{dequeue} operations return the
smallest/largest \emph{enqueue}d value which has not yet been dequeued. It
appears that such ADTs are rare in practice. Overcoming this limitation would
require a refinement to matching schemes which can track additional relations
among data values.

>>>>

First, while our theoretical foundation does cover “concurrency-aware” ADT
specifications~\cite{conf/podc/HemedR14}, i.e.,~non-atomic ADTs like the
\emph{rendezvous synchronizer}, \emph{barrier}, and \emph{exchanger}, which
have bounded width greater than $1$, our current implementation only
handles atomic ADTs. In principle, this limitation is not fundamental. The key
difference is in the enumeration ADT histories. For a given ADT,
determining whether a given history is admitted or not reduces
to checking whether there exists an execution of which the history is an
abstraction. For atomic ADTs, only a single sequential execution need be
examined, in which no two operations overlap. For non-atomic ADTs, every
possible interleaving of the internal actions of operations need be examined.
While we expect this calculation to remain feasible given that ADT complements
are normally represented with histories with few operations, we do expect it to
incur a noticeable cost.

Second, ADTs whose specifications rely on relations (besides equality) on
method argument and return values cannot be expressed with our notion of
histories based on matching schemes. A notable example of such an ADT is the
\emph{priority queue}, whose \emph{dequeue} operations return the
smallest/largest \emph{enqueue}d value which has not yet been dequeued.
%It appears that such ADTs are rare in practice.
Overcoming this limitation would
require a refinement to matching schemes which can track additional relations
among data values.

### 2015-07-09 formula.tex
Removed kernel talk.

While Section~\ref{sec:patterns} demonstrates that (the kernel complements) of
naturally-occurring ADTs can be represented finitely, in this section we
demonstrate that such finite representations have logical interpretations,
allowing us to derive the formulas representing ADTs. In what follows, we
describe how to obtain a formula which is satisfied by the histories which
embed a given history, via $\preceq$. Then, using the finite set of histories
which represent the complement of a given ADT, we represent the ADT itself as
the conjunction of negations of these embedding formulas. The resulting formula
is satisfied only on the histories which do not embed the generators of a given
ADT’s complement.

>>>>

While Section~\ref{sec:patterns} demonstrates that (the complements) of
naturally-occurring ADTs can be represented finitely, in this section we
demonstrate that such finite representations have logical interpretations,
allowing us to derive the formulas representing ADTs. In what follows, we
describe how to obtain a formula which is satisfied by the histories which
embed a given history, via $\preceq$. Then, using the finite set of histories
which represent the complement of a given ADT, we represent the ADT itself as
the conjunction of negations of these embedding formulas. The resulting formula
is satisfied only on the histories which do not embed the generators of a given
ADT’s complement.

### 2015-07-09 nature.tex
Fixed space hacks.
Section titles + restructuring the restructuring.
Removed kernel talk.

%!TEX root = draft.tex
\section{Properties of Naturally-Occurring ADTs}
\label{sec:nature}

\makeatletter
\renewcommand{\verbatim@font}{\ttfamily\scriptsize}
\makeatother

In this section we demonstrate that the premises used in the development of our
symbolic ADT inference algorithm — i.e.,~unique and bounded-width kernels,
faithful and normalizing matching schemes, and predictability — hold for the
ADTs which are typically provided by concurrent object libraries. Furthermore,
we demonstrate that the algorithm developed in this work computes precise
symbolic representations for these ADTs which can be used in symbolic checkers
for observational refinement~\cite{conf/pldi/EmmiEH15}. Our publicly available
implementation\footnote{The URL has been suppressed for anonymity purposes.}
enumerates the sequential histories\footnote{Our current implementation is
limited to atomic ADTs. For non-atomic ADTs, the enumeration must cover all
$k$-width histories, for some $k \in \mathbb{N}$.} not admitted by reference
implementations of the undermentioned ADTs, keeping only the histories which
are not generated by any other. In each case, our algorithm terminates in a
matter of seconds.

@@

\end{itemize}
As its name implies, its kernel is sequential.

@@

It is easy to see that the atomic register is normal. It is
closed under removal of read-only and duplicate operations, since  each
match can contain an arbitrary number of read operations. Since the histories
of atomic registers do not contain unmatched operations, they are also closed
under their removal. Finally, since the kernel contains an arbitrary number of
sequenced matches, atomic register histories are also closed under match
removal.

@@

\end{itemize}
As their names imply, their kernels are sequential.

@@

\end{itemize}
Each operation returns the nil value $v = -$ when $u$ is not yet
present, and otherwise returns $v = u$. As its name implies, its kernel is
sequential.

@@

\end{itemize}
Unlike atomic queues, the work-stealing queue permits values to be removed
twice: once normally, via the take operation, and once exceptionally, via the
steal operation. We give a faithful matching scheme over executions which add
unique values analogously to the scheme for stacks and queues, in which give
operations match themselves, while take and steal operations match the give
operation which added their returned value, or themselves, in case the nil
value is returned. Note that the kernel of work-stealing queues has width $2$,
since a pairs of concurrent take and steal operations returning the same value
are permitted, while sequentially they are not. As our implementation is
currently limited to width-$1$ histories (see the discussion in
Section~\ref{sec:discussion}), below we generate only the width-$1$ complement.

>>>>

%!TEX root = draft.tex
\section{Naturally-Occurring ADTs}
\label{sec:nature}

In this section we demonstrate that the premises used in the development of our
symbolic ADT inference algorithm — i.e.,~uniqueness and bounded-width,
faithful and normalizing matching schemes, and predictability — hold for the
ADTs which are typically provided by concurrent object libraries. Furthermore,
we demonstrate that the algorithm developed in this work computes precise
symbolic representations for these ADTs which can be used in symbolic checkers
for observational refinement~\cite{conf/pldi/EmmiEH15}. Our publicly available
implementation\footnote{The URL has been suppressed for anonymity purposes.}
enumerates the sequential histories\footnote{Our current implementation is
limited to atomic ADTs. For non-atomic ADTs, the enumeration must cover all
$k$-width histories, for some $k \in \mathbb{N}$.} not admitted by reference
implementations of the undermentioned ADTs, keeping only the histories which
are not generated by any other. In each case, our algorithm terminates in a
matter of seconds.

@@

\end{itemize}
As its name implies, the atomic register ADT is sequential.

@@

It is easy to see that the atomic register is normal. It is
closed under removal of read-only and duplicate operations, since  each
match can contain an arbitrary number of read operations. Since the histories
of atomic registers do not contain unmatched operations, they are also closed
under their removal. Finally, since the sequences contain an arbitrary number of
matches, atomic register histories are also closed under match
removal.

@@

\end{itemize}
As their names imply, these ADTs are sequential.

@@

\end{itemize}
Each operation returns the nil value $v = -$ when $u$ is not yet
present, and otherwise returns $v = u$. As its name implies, the atomic set
ADT is sequential.

@@

\end{itemize}
Unlike atomic queues, the work-stealing queue permits values to be removed
twice: once normally, via the take operation, and once exceptionally, via the
steal operation. We give a faithful matching scheme over executions which add
unique values analogously to the scheme for stacks and queues, in which give
operations match themselves, while take and steal operations match the give
operation which added their returned value, or themselves, in case the nil
value is returned. Note that the work-stealing queue ADT has width $2$,
since a pairs of concurrent take and steal operations returning the same value
are permitted, while sequentially they are not. As our implementation is
currently limited to width-$1$ histories (see the discussion in
Section~\ref{sec:discussion}), below we generate only the width-$1$ complement.

### 2015-07-09 related.tex
Removed kernel talk.

The idea of decomposing ADT violations into a finite set of patterns has been
proposed for atomic collections~\cite{conf/tacas/AbdullaHHJR13,
conf/concur/HenzingerSV13, conf/popl/DoddsHK15, conf/popl/BouajjaniEEH15}, and
generalized to ADTs which can be expressed with a particular form of recursive
definition~\cite{journals/arxiv/BouajjaniEEH15}. These works suggest patterns
which directly recognize violations in \emph{concurrent} executions,
effectively reducing observational refinement to the verification of classical
temporal-logic properties. Some of these patterns are expressible only as
\emph{infinite}, though regular, languages. On the contrary, in this work we
suggest patterns expressed as \emph{finite} languages, which recognize
violations in ADT \emph{kernels} — i.e.,~the sequential executions of atomic
objects, or the “concurrency-aware” executions of “concurrency-aware”
specifications~\cite{conf/podc/HemedR14}. Though by themselves these patterns
are incomplete in recognizing violations in concurrent executions, they are
complete when considered as negative conditions on the linearizations of
concurrent executions, effectively reducing verification of a single execution
to satisfiability checking~\cite{conf/pldi/EmmiEH15}.

>>>>

The idea of decomposing ADT violations into a finite set of patterns has been
proposed for atomic collections~\cite{conf/tacas/AbdullaHHJR13,
conf/concur/HenzingerSV13, conf/popl/DoddsHK15, conf/popl/BouajjaniEEH15}, and
generalized to ADTs which can be expressed with a particular form of recursive
definition~\cite{journals/arxiv/BouajjaniEEH15}. These works suggest patterns
which directly recognize violations in \emph{concurrent} executions,
effectively reducing observational refinement to the verification of classical
temporal-logic properties. Some of these patterns are expressible only as
\emph{infinite}, though regular, languages. On the contrary, in this work we
suggest patterns expressed as \emph{finite} languages, which recognize
violations in ADTs — i.e.,~the sequential executions of atomic
objects, or the “concurrency-aware” executions of “concurrency-aware”
specifications~\cite{conf/podc/HemedR14}. Though by themselves these patterns
are incomplete in recognizing violations in concurrent executions, they are
complete when considered as negative conditions on the linearizations of
concurrent executions, effectively reducing verification of a single execution
to satisfiability checking~\cite{conf/pldi/EmmiEH15}.
